\pkg{plotROC}: A Better Tool for Plotting ROC Curves

### 2015-10-14 Paper/response-to-review.md

# Reviewer 1

> This is a really cool package that uses state of the art tech with ggplot and shiny, as well as interactive plots, to display ROC curves.
> Unfortunately, the package is rather difficult to use and quite inconsistent. It also produces some bogus results (see below). As it is now, it looks more like a pretty cool tech demo than a scientific software I would use in my own research. Here are a few comments to try to improve it, as well as the manuscript itself.

## Bug report:
> The confidence intervals are apparently bogus. The following code executed in Rstudio or from the command line, with R 3.1.3 on Ubuntu 14.04 (from CRAN's Debian repository) and R 3.1.2 on Mac OS Yosemite (from Fink):

set.seed(42)
D.ex <- rbinom(100, size = 1, prob = .5)
M.ex <- rnorm(100, mean = D.ex)
roc.estimate <- calculate_roc(round(M.ex, 2), D.ex, ci = TRUE)
plot_interactive_roc(ggroc(roc.estimate, ci = TRUE))

After clicking on a point around 1.6, I got the screenshot shown in Figure 1. This seems related to the rounding (here to the 2nd decimal).
Figure 1: the CI region doesn't span the cutoff point.

Thank you for the bug report. I believe this was due to the way the previous function handled ties in the results. The computation of the ROC curve in R and the js functions for interactivity have been completely revamped. I cannot reproduce the error using the new update in any browser that I've tested, so I believe it has been fixed.

## General:
> • The manuscript is written partially in the first person. It is surprising and not totally consistent throughout the manuscript, some of which is written in passive tense (for instance “Labels can be added easily [...]”).

The manuscript has been edited to be consistent with the use of active voice.

> • The abstract highlights the literature review as a significant finding. Surprisingly it is is only briefly mentioned in the introduction. This seems a bit inconsistent.

> • Strictly speaking one could argue the plotROC package "obscures the cutoffs", as they are not displayed by default when calling the ggroc function. This is not so different from existing packages (as reviewed in subsection 1.3) that allow accessing the thresholds and adding them to the plot in different ways. The user has to request an interactive or journal plot to get them. The abstract and introduction should be rephrased in a more neutral way on this topic, as this is not the key point addressed by this package.
## Introduction:
> • The introduction states that "ROC curves have been used for decades in signal detection theory". In fact, they originate from this field (s/n in radio receivers, hence the name)
> • The "good outcome", "bad outcome" is confusing to design correct / incorrect classifications. Especially in the medical field. I don't think it is necessary.
> • "A perfect test [...] has TPF(c) = 1 and FPF(c) = 0 for all c". This is not correct. It is sufficient that this equality holds true for a single c.
> • The formula hat(TPF(c)) is slightly confusing. It may be useful to first introduce the confusion table with true | false positive | negatives. It seems correct nonetheless.
> • "The greater the AUC, the more informative the test". This statement is incomplete at best. What matters is | AUC - 0.5 |. An AUC = 0 is also the result of a perfect test (for which M < c if I am not mistaken).
> • Existing software: pROC allows to plot cutoffs quite easily (with print.thres=TRUE, just checked now). Of course it still uses the base R plots.
> • I am surprised a Pubmed search returned only 54 papers. This part would probably deserve to be expanded a little bit, and maybe turned into a whole section.
## Usage of the package:
> • Plots such as fig 4 will be hard to read in small figures, as is often seen in journals.
> • Fig 5 is unclear. Especially, what was the cutoff used to compute the region, as several ones are displayed in the region?
> • In general, the use of shaded region instead of bars is quite surprising, although perhaps more accurate than bars.
> • Why are the regions rectangle-shaped and not oval? That would seem even more correct, although probably quite difficult to set up.
> • The gridlines in figures 3-5 & 7 are so thin and light-coloured that they did not appear when I printed this manuscript on paper. The printer was not (yet) running out of ink.
> • When reporting an AUC and confidence intervals in a publication, it seems to me at least as important to report values in addition to a plot. One can annotate a curve with the AUC (p. 13), with, for instance, label = “AUC=0.80”. But this value comes out of nowhere, and does not even look realistic (to me it looks more like AUC = 0.7). In addition, it isn't an “Advanced option”, as the subsection suggests, but rather as something quite basic. An easy way out would be to add a few additional lines computing this AUC to the manuscript. The case of CI seem a bit more difficult to deal with.

## Package:

> • The usage requires a lot of typing, and does not appear to be straightforward. For instance, the function calculate_roc must be used to create the FPF and TPF matrix. This will create a lot of headaches for infrequent users: is it calculate_, or create_, or generate_, or ...?
> • This wouldn't be an issue if users could browse the functions of the package as one can usually do by typing ?plotROC. Unfortunately, this leads to a stub that has never been authored. Please update man/plotROC-package.R with some useful pointers. Of course one can always find this paper, but it is much easier to get a proper help.
> • It is of course a design choice, but the use of underscores over dots (or camelCase) in function names seems quite surprising and may disturb some users. See Bååth, R. (2012). The State of Naming Conventions in R. R Journal, 4, 74-75. R Foundation for Statistical Computing for a review of current usage of function name styles.
> • Wouldn't it be more idiomatic to introduce the roc as a ggplot geometry? That would allow some much more straightforward syntax like:
data.for.roc <- data.frame(M.ex, D.ex)
qplot(M.ex, D.ex, data = data.for.roc) + geom_roc()
or suchlike that would be much easier to remember.

> • I don't understand why one has to pass a ggroc argument to plot_interactive_roc. That is a lot of functions to use in row and in the right order to get an interactive plot, and 3 times (_)?roc. I suspect a few well designed objects would make it much more user-friendly.
> • What do the modCss and modJs functions do? The documentation is really limited and an example would help. Same with verify_d. As it stands now these functions seem useless.
> • The help files lack examples altogether. Not only does it help a lot to have copy- pasteable code snippets as a start, but it can also help checking the package for potential regressions during updates.
> • Performance: as most R packages doing ROC curves, this one scales pretty badly with larger sample sizes. I recommend using the cumsum function to compute the FPF/TPF. This will be much faster than the current implementation with sapply (which is really just a hidden slow loop). See the ROCR package that does it correctly. The calculate_roc function with the sample data of subsection 2.4 scaled to 1E5 data points literally collapsed and took > 10 minutes to compute. Such sample sizes may be uncommon in oncology, but are not rare in genomics or computer science. ROCR returns nearly instantly with such a low (sic!) sample size and scales much higher.
> • How do I add the cut-offs to the plot? If I say ggplot(..., ci.at = 0.5) I get an unused argument error and nothing is added. Is this only available in the interactive plot?

## Shiny app:

> • The shiny application has an hard dependency on ggthemes and is unusable without that package. Unfortunately it is only “Suggest”ed, and and error message is printed only after a query to the shiny web application that doesn't work. PlotROC should either “Depend” on that package, as for shiny, or catch its absence and print the error message immediately when the shiny app is launched.
> • On my Ubuntu system, the Download button opens a 404 error page with URL as http://127.0.0.1:3048/session/61edd0aac13532dc34788b220e9d27c8/downlo ad/printDownload?w= instead of downloading the image. I can right click the image and “View image” normally. The download works properly on my Mac where a tab shortly opens before the download starts.
> • The browse button appears very low in the page and is not visible enough IMO.

# Reviewer 2

> The paper about plotROC describes a new R package for plotting receiver operating characteristic curves (ROC) curves for the evaluation of binary classifiers.

> The paper is well written and gives background on ROC curves, motivation for improving ROC curves based on a literature survey, as well as code examples on how to use plotROC.

> The greatest strengths of the new package are: 
	- Clean and nice graphics (based on ggplot2)
	- The attempt to provide sensible defaults (e.g. for fontsize, axis tickmarks, etc.)
	- Options for interactivity, i.e. providing a web-interface using shiny. This is intended to make high quality ROC analyses accessible to non-technical users

> However, the package falls short of the goal proclaimed in the title of providing a "better tool for plotting ROC curves". While the strengths above deserve credit, other established tools (e.g. the referenced packages pROC and ROCR) provide several other features that are not covered in the current package (e.g. other performance measures, AUC, partial AUC, averaging of curves, aggregating curves across bootstraps or cross-validation folds, and confidence limits and regions on many of those measures).

> Also the paper and package suffer from a number of ambiguities and inaccuracies (either technical or in the description):
	- The extreme point TPF/TPR=1/1 is not included in the output of the calculate_roc function. This could lead to problems when plotting curves that don't span the full 0/0 to 1/1 space and also when computing AUCs.
	- plotROC seems to implictly define that samples ">" the threshold are positive, samples "<=" the threshold are negative. In this it differs from e.g. Fawcett "An introduction to ROC analysis" or other software (e.g. the ROCR package) that define samples ">=" the threshold as a positive prediction.
	- It is not obvious how plotROC handles ties (i.e. samples with identical numerical measurements)
	- The paper defines "good outcome" and "bad outcome" in Section 1.1. These definitions seem to ignore half of the confusion matrix (i.e. ignoring true and false negatives).
	- In "Help and Documentation" of the webpage, the confusion matrix and the definitions seem to be copied from Wikipedia. Credits and reference are needed.
	- The website states "confidence level = 0.05". Should this be "significance level" instead?

> Overall, the paper and package provide some new noteworthy ideas to ROC analysis with the attempt to make high quality analyses more accessible to non-technical experts. The current tool does not have the maturity and comprehensiveness of other tools - therefore the users will be left with the choice between alternative tools, each with their own strengths and weaknesses.

### 2015-10-14 Paper/sachs2015plotROC.Rmd
compiled

opts_chunk$set(dev="tikz", prompt = TRUE)

@@

\begin{table}
\begin{tabular}{l|l|c|c|c}
\multicolumn{2}{c}{}&\multicolumn{2}{c}{$D$}&\\
\cline{3-4}
\multicolumn{2}{c|}{}&1&0&\multicolumn{1}{c}{Total}\\
\cline{2-4}
\multirow{2}{*}{$M$}& $M \geq c$ & true positive & false positive & test positive\\
\cline{2-4}
& $M < c$ & false negative & true negative & test negative\\
\cline{2-4}
\multicolumn{1}{c}{} & \multicolumn{1}{c}{Total} & \multicolumn{1}{c}{disease positive} & \multicolumn{1}{c}{disease negative} & \multicolumn{1}{c}{total}\\
\end{tabular}
\caption{\label{confus} Confusion matrix illustrating the possible classification outcomes of a continuous test $M$ with threshold $c$ versus the true outcome $D$. }
\end{table}

@@

There are several \proglang{R} packages related to ROC curve estimation that contain dedicated plotting functions. The \pkg{ROCR} package [@rocr] plots the FPF _versus_ TPF, as usual, and then takes the interesting approach of encoding the cutoff values as a separate color scale along the ROC curve itself. A legend for the color scale is placed along the vertical axis on the right of the plotting region. The \pkg{pROC} package [@pROC] provides an option for plotting cutoff labels (`print.thres = TREU`) and is mainly focused on estimating confidence intervals and regions for restricted ranges of the ROC curve. The plotting methods therein use the base \proglang{R} plotting functions to create nice displays of the curves along with shaded confidence regions. My \pkg{plotROC} package uses the \pkg{ggplot2} [@ggplot2] plotting library to create clear, informative ROC plots, with interactive features for use on the web, and sensible defaults for use in print.

@@

```{r figure1, fig.width = 6, fig.height = 6, echo = FALSE, fig.cap = "Illustration of design choices in plotting ROC curves. Panel A shows a sparse ROC curve, with no design additions inside the plotting region. The plot results in more white space than anything else. It is difficult to accurately determine values without reference lines. Panel B shows a plot comparing 2 curves, with different line types and a legend. AUCs are also given in the legend. Panels C and D add gridlines, diagonal reference lines, and direct labels. \\label{figure1}"}
library(plotROC)

@@

```{r test-a-ci, eval = 1, fig.cap = "Illustration of \\pkg{plotROC} with exact confidence regions \\label{ciex}."}
basicplot + geom_rocci()

@@

```{r print2, warning = FALSE, message = FALSE, fig.width = 6, fig.height = 6, fig.cap = "Using \\pkg{ggplot2} themes and annotations with \\pkg{plotROC} objects. \\label{annotate}"}
basicplot + 
  style_roc(theme = theme_grey) +
  theme(axis.text = element_text(colour = "blue")) +
  ggtitle("Themes and annotations") + 
  annotate("text", x = .75, y = .25, 
           label = paste("AUC =", round(calc_auc(basicplot)$AUC, 2))) +
  scale_x_continuous("1 - Specificity", breaks = seq(0, 1, by = .1))
```

@@

There are many ways to solve this with \pkg{d3.js}, but I decided to use Voronoi polygons to map the cursor location to the nearest point on the ROC curve. The idea is that for the set of cutoff points along the ROC curve, the \code{d3.geom.voronoi} function chain computes a set of polygons overlaying the plotting region such that the area of each polygon contains the region of the plot closest to it's corresponding cutoff point. Hover events are bound to the polygons so that when the mouse cursor moves around the plotting region, the closest point on the ROC curve is made visible. Similarly, click events are bound to the polygons so that the appropriate confidence region is made visible upon clicking. The svg code and all necessary \proglang{JavaScript} code is returned in the character string provided by \code{export_interactive_roc}. Figure \ref{flow} outlines the \pkg{plotROC} process.

```{r diagram, include = FALSE}
library(DiagrammeR)

DiagrammeR("
  graph LR;
    A[ROC Estimate]-->|ggroc|B(ggplot object);
    B-->|plot_journal_roc|C(Static Plot);
    B-->|grid.export|D(svg object);
    D-->E(Interactive Plot);
    F(d3.js functions)-->E;
    G(css styles)-->E;

  style A fill:#DCEBE3
  style C fill:#A2EB86
  style E fill:#A2EB86
")

## export as pdf

\begin{figure}[ht]
\centering
\includegraphics{figure/diagram.pdf}
\caption{Flowchart illustrating the approach that \pkg{plotROC} takes to generate either static plots for print or interactive plots for web-use. \label{flow}}
\end{figure}

>>>>

opts_chunk$set(dev="tikz", prompt = TRUE, sanitize = TRUE)

@@

\begin{table}
\begin{tabular}{l|c|c|c}
\multicolumn{1}{c}{}&\multicolumn{2}{c}{$D$}&\\
\cline{2-3}
\multicolumn{1}{c|}{}&1&0&\multicolumn{1}{c}{Total}\\
\cline{2-3}
 $M \geq c$ & true positive & false positive & test positive\\
\cline{2-3}
 $M < c$ & false negative & true negative & test negative\\
\cline{2-3}
\multicolumn{1}{c}{Total} & \multicolumn{1}{c}{disease positive} & \multicolumn{1}{c}{disease negative} & \multicolumn{1}{c}{}\\
\end{tabular}
\caption{\label{confus} Confusion matrix illustrating the possible classification outcomes of a continuous test $M$ with threshold $c$ versus the true outcome $D$. }
\end{table}

@@

There are several \proglang{R} packages related to ROC curve estimation that contain dedicated plotting functions. The \pkg{ROCR} package [@rocr] plots the FPF _versus_ TPF, as usual, and then takes the interesting approach of encoding the cutoff values as a separate color scale along the ROC curve itself. A legend for the color scale is placed along the vertical axis on the right of the plotting region. The \pkg{pROC} package [@pROC] provides an option for plotting cutoff labels (\code{print.thres = TRUE}) and is mainly focused on estimating confidence intervals and regions for restricted ranges of the ROC curve. The plotting methods therein use the base \proglang{R} plotting functions to create nice displays of the curves along with shaded confidence regions. My \pkg{plotROC} package uses the \pkg{ggplot2} [@ggplot2] plotting library to create clear, informative ROC plots, with interactive features for use on the web, and sensible defaults for use in print.

@@

```{r figure1, fig.width = 6, fig.height = 6, message = FALSE, echo = FALSE, fig.cap = "Illustration of design choices in plotting ROC curves. Panel A shows a sparse ROC curve, with no design additions inside the plotting region. The plot results in more white space than anything else. It is difficult to accurately determine values without reference lines. Panel B shows a plot comparing 2 curves, with different line types and a legend. AUCs are also given in the legend. Panels C and D add gridlines, diagonal reference lines, and direct labels. \\label{figure1}"}
library(plotROC)

@@

```{r test-a-ci, eval = 1, fig.cap = "Illustration of \\pkg{plotROC} with exact confidence regions. \\label{ciex}"}
basicplot + geom_rocci()

@@

```{r print2s, warning = FALSE, message = FALSE}
annotate <- basicplot + style_roc(theme = theme_grey) +
  theme(axis.text = element_text(colour = "blue")) +
  ggtitle("Themes and annotations") + 
  annotate("text", x = .75, y = .25, 
           label = paste("AUC =", round(calc_auc(basicplot)["AUC"], 2))) +
  scale_x_continuous("1 - Specificity", breaks = seq(0, 1, by = .1))
```

```{r print2, echo = FALSE, fig.width = 6, fig.height = 6, fig.cap = "Using \\pkg{ggplot2} themes and annotations with \\pkg{plotROC} objects. \\label{annotate}"}
annotate
```

The results are compatible with all other \pkg{ggplot2} layers and functions, and interactive versions are supported as long as there is an Roc layer present.

@@

There are many ways to solve this with \pkg{d3.js}, but I decided to use Voronoi polygons to map the cursor location to the nearest point on the ROC curve. The idea is that for the set of cutoff points along the ROC curve, the \code{d3.geom.voronoi} function chain computes a set of polygons overlaying the plotting region such that the area of each polygon contains the region of the plot closest to it's corresponding cutoff point. Hover events are bound to the polygons so that when the mouse cursor moves around the plotting region, the closest point on the ROC curve is made visible. Similarly, click events are bound to the polygons so that the appropriate confidence region is made visible upon clicking. The svg code and all necessary \proglang{JavaScript} code is returned in the character string provided by \code{export_interactive_roc}.
